\documentclass[12pt]{article}

\begin{document}

{\LARGE Final Project }


\noindent 1. Generate random $[x,y,\delta y]$ data points using the following code

\begin{verbatim}
data = [(x,random.gauss(1,0.01)*exp(log(2)+log(id)*x/400),20) 
        for x in range(100,201)]
\end{verbatim}

\noindent where {\tt id } is your student id.

\noindent 2. Plot the data.

\noindent 3. The data can be modeled with the following function

\[
y = 2 e^{b x}
\]

\noindent You must determine the value of $b$ in three different ways:

\begin{itemize}
\item Analitycally, as function of your student id.
\item Using the linear least square fitting function discussed in class. Notice the fitting function above is not linear in $b$ therefore you need to perform a transformation of the data before you can apply linear least squares.
\item From first principles, by minimizing $\chi^2$ using an optimization algorithm (for example Newton).
\end{itemize}

For each of the above methods:
\begin{itemize}
\item Explain the method in detail
\item Show your steps
\item Show your result
\item A plot of your fitting function superimposed to the data
\end{itemize}

\noindent Do you expect the three methods to agree? Why? If you do not expect them to agree exactely, to what precision should they agree?

{\bf Deliverables}

You must deliver a 5 pages paper in PDF (Word documents will not be accepted) solving the problem described above. The paper should include an introduction explaining the problem in your own language. Imagine the reader of your paper your boss at work and not somboby like me who alreday understands the problem. The paper should include one section for each of the three methods described above. The paper should include 4 plots as requested above. The paper should have a conclusion answering the above questions.

The paper should have an appending listing all your code. The code must be properly formatted/indented.

Be aware that this is an individual project. The data should be different for every student and different papers will be checked for similarities. Excessive similarities will be reported for plagiarism.

\end{document}
