\documentclass[12pt]{article}

\usepackage{graphicx}
\usepackage{url}
\usepackage{upquote}
\begin{document}

\section{CSC299 - Fall 2013 - Lab Assignment 9}

\subsection*{IMPORTANT INSTRUCTIONS}

\noindent Login into:
\begin{verbatim}
mdp.cdm.depaul.edu
\end{verbatim}

\noindent In this class you will use {\tt fabric} to authomate some tasks.
\begin{verbatim}
http://docs.fabfile.org/
\end{verbatim}

\noindent Under your csc299 folder create a new folder called {\tt lab09}.

\noindent After completing each task below remember to copy your {\tt fabfile} under {\tt lab8} and commit it to github.

\noindent If your account is {\tt user<nnn>} than I have created another account for you {\tt user<nnn>tmp} with password {\tt 987654321}. You will run locally using account {\tt user<nnn>} but you will run tasks on {\tt user<nnn>tmp@mdp.cdm.depaul.edu} using fabric. 

\noindent Your new account contains a folder {\tt /home/user<nnn>/web2py}. You do not need to know what it contains. All you need to know is that it is a folder which contains other files and subfolders.

\noindent Your programs should use the {\tt settings} context manager to handle failures.

\subsection{Task 1 (2 points)}

Use the command {\tt git rm <filename>} to remove all unwanted files from your repository (tests, files starting with a . or a \verb!#! or ending in \verb!~!. Commit your changes.

\subsection{Task 2 (3 points)}

Create a ssh key in order to be able to login to {\tt user<nnn>tmp@mdp.cdm.depaul.edu} without need for a password.
Add and commit a copy of your public key in the {\tt lab09} folder.

\subsection{Task 3 (6 points)}

Build a fabric script {\tt lab09/fabfile.py} that uses tha above certificate for login. The task of the fabric script is that of downoloading your Bitbucker git repository (using git clone) for the class in the new account and runs all the programs you have created so far (labXXX/programYYY.py) on the new account. The fabric script should also zip the remote {\tt csc299} folder and retrieve it. Commit the fabric script. The fabric script itself should not perform {\tt git} operations other than {\tt git clone}.

\end{document}
