\documentclass[12pt]{article}

\usepackage{graphicx}
\usepackage{url}
\usepackage{upquote}
\begin{document}

\section{CSC299 - Fall 2015 - Lab Assignment 7}

\subsection*{IMPORTANT INSTRUCTIONS}

\noindent Login into:
\begin{verbatim}
mdp.cdm.depaul.edu
\end{verbatim}

\noindent Under your csc299 folder create a new folder called {\tt lab07}.
\noindent After completing each task below remember to do:

\begin{verbatim}
git add README.json
git add .
git commit -a -m "task completed"
git push
\end{verbatim}

\noindent  else I will not receive your work.

\noindent In this class you MUST use mechanize and must connect using a csc299 header, as in the following example:

\begin{verbatim}
import mechanize
br = mechanize.Browser()
br.addheaders = [('csc299','true')]
br.open('http://mdp.cdm.depaul.edu/csc299a7/default/index')
br.forms()
\end{verbatim}

\subsection{Task 1 (3 points)}

\noindent  In folder {\tt ~/csc299/lab07} write a program {\tt program71.py} that connects to 

\begin{verbatim}
http://mdp.cdm.depaul.edu/csc299a7/default/user/register
\end{verbatim}

and registers you. Use your actual email and student id.
DO NOT REGISTER USING THE BROWSER ELSE YOU WILL NOT BE ABLE TO REGISTER TWICE.
After you register using mechanize, you can try it using the browser using the email and password provided upon registration.

\subsection{Task 2 (3 points)}

\noindent Write a program {\tt program72.py} that connects to 

\begin{verbatim}
http://mdp.cdm.depaul.edu/csc299a7/default/index
\end{verbatim}

and does the following:
\begin{itemize}
\item Sign in
\item Extract all numbers from the {\tt index} page.
\end{itemize}

Notice the sequence of numbers is different for each students and randomized at each call to the {\tt index} page.

\subsection{Task 3 (3 points)}

\noindent Modify the above program in order to:
\begin{itemize}
\item Sign in
\item Extract all numbers from the {\tt index} page.
\item Compute the sum of the numbers ({\tt total})
\item Click the ``Next'' button
\end{itemize}

\subsection{Task 4 (2 points)}

\noindent Modify the above program in order to:
\begin{itemize}
\item Sign in
\item Extract all numbers from the {\tt index} page.
\item Compute the sum of the numbers ({\tt total})
\item Click the ``Next'' button
\item In the next page ({\tt form}), submit the total
\item Check that the page contains ``Assignment Completed''
\end{itemize}

\subsection*{Useful links}

\begin{verbatim}
http://mdp.cdm.depaul.edu/csc299a7/default/user/register
http://mdp.cdm.depaul.edu/csc299a7/default/user/logout
http://mdp.cdm.depaul.edu/csc299a7/default/user/login                     
http://mdp.cdm.depaul.edu/csc299a7/default/user/index
http://mdp.cdm.depaul.edu/csc299a7/default/user/form
http://mdp.cdm.depaul.edu/csc299a7/default/user/done
\end{verbatim}

Note you only need to open the ``register'' and the ``index'' and perhaps the ``logout'' link. Everything is handled by clicking and form redirections automatically.

\end{document}
