\documentclass[12pt]{article}

\usepackage{graphicx}
\usepackage{url}
\usepackage{upquote}
\begin{document}

\section{CSC299 - Fall 2013 - Lab Assignment 8}

\subsection*{IMPORTANT INSTRUCTIONS}

\noindent In this class you will use {\tt fabric} to authomate some tasks.
\begin{verbatim}
http://docs.fabfile.org/
\end{verbatim}

\noindent In this assignment you will have to use two accounts:
\begin{verbatim}
csc299-<student-id>@dpu.experts4solutions.com
csc299-<student-id>@mdp.cdm.depaul.edu
\end{verbatim}
The password is the same.

You will login into the former (dpu) account and using fabric perform remote tasks into the latter account (mdp).

\subsection{Task 1 (3 points)}

Edit the fabfile and create a task that copies fabfile itself into the remote folder 
{\verb!mdp.cdm.depaul.edu:~/csc299/lab09/!}
Deploy you fabfile.

\subsection{Task 2 (3 points)}

Edit the fabfile and create a task that zips your remote {\verb!mdp.cdm.depaul.edu:~/csc299!} folder and retrieves the zip file.

After done, copy your fabfile again into {\verb!mdp.cdm.depaul.edu:~/csc299/lab09/!}.

\subsection{Task 3 (4 points)}

\noindent In task three you will dowload the latest web2py locally, install it remotely, and run it.

\noindent Amend the fabfile to provide the following command:
\begin{verbatim}
fab install_web2py
\end{verbatim}
\noindent The corresponding task should perform the following steps:
\begin{itemize}
\item locally download
{\footnotesize\begin{verbatim}
http://www.web2py.com/examples/static/web2py_src.zip
\end{verbatim}}
\item deploy the file remotely and unzip it in your home folder.
\item remotely under {\tt web2py} run 
\begin{verbatim}
python web2py.py -i 0.0.0.0 -p <port> -a <pwd>
\end{verbatim} where {\tt <port>} is the digit 9 followed by the last 3 digits of your student id. {\tt <pwd>} is a secure password of your choosing.
\end{itemize}
\noindent From the browser try open 
{\footnotesize\begin{verbatim}
http://mdp.cdm.depaul.edu:<port>/welcome
\end{verbatim}}  
\noindent to check you have successfully installed web2py.

After done, copy your fabfile again into {\verb!mdp.cdm.depaul.edu:~/csc299/lab09/!}.

\end{document}
